---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# compositional.mle

<!-- badges: start -->
[![R-CMD-check](https://github.com/queelius/compositional.mle/workflows/R-CMD-check/badge.svg)](https://github.com/queelius/compositional.mle/actions)
<!-- badges: end -->

An R package for **composable maximum likelihood estimation**. Solvers are first-class functions that combine via sequential chaining, parallel racing, and random restarts.

## Installation

```r
# install.packages("devtools")
devtools::install_github("queelius/compositional.mle")
```
## Design Philosophy

Following SICP principles, the package provides:

1. **Primitive solvers** - `gradient_ascent()`, `newton_raphson()`, `bfgs()`, `nelder_mead()`, etc.
2. **Composition operators** - `%>>%` (sequential), `%|%` (race), `with_restarts()`
3. **Closure property** - Combining solvers yields a solver

## Quick Start

```{r example, message=FALSE}
library(compositional.mle)

# Generate sample data
set.seed(42)
x <- rnorm(100, mean = 5, sd = 2)

# Define the problem (separate from solver strategy)
problem <- mle_problem(
  loglike = function(theta) {
    if (theta[2] <= 0) return(-Inf)
    sum(dnorm(x, theta[1], theta[2], log = TRUE))
  },
  score = function(theta) {
    mu <- theta[1]; sigma <- theta[2]; n <- length(x)
    c(sum(x - mu) / sigma^2,
      -n / sigma + sum((x - mu)^2) / sigma^3)
  },
  constraint = mle_constraint(
    support = function(theta) theta[2] > 0,
    project = function(theta) c(theta[1], max(theta[2], 1e-8))
  )
)

# Simple solve
result <- gradient_ascent()(problem, theta0 = c(0, 1))
result$theta.hat
```

## Composing Solvers

### Sequential Chaining (`%>>%`)

Chain solvers for coarse-to-fine optimization:

```{r sequential}
# Grid search -> gradient ascent -> Newton-Raphson
strategy <- grid_search(lower = c(-10, 0.5), upper = c(10, 5), n = 5) %>>%
  gradient_ascent(max_iter = 50) %>>%
  newton_raphson(max_iter = 20)

result <- strategy(problem, theta0 = c(0, 1))
result$theta.hat
```

### Parallel Racing (`%|%`)

Race multiple methods, keep the best:

```{r race}
# Try multiple approaches, pick winner by log-likelihood
strategy <- gradient_ascent() %|% bfgs() %|% nelder_mead()

result <- strategy(problem, theta0 = c(0, 1))
c(result$theta.hat, loglike = result$loglike)
```

### Random Restarts

Escape local optima with multiple starting points:

```{r restarts}
strategy <- with_restarts(
  gradient_ascent(),
  n = 10,
  sampler = uniform_sampler(c(-10, 0.5), c(10, 5))
)

result <- strategy(problem, theta0 = c(0, 1))
result$theta.hat
```

## Available Solvers

| Factory | Method | Requires |
|---------|--------|----------|
| `gradient_ascent()` | Steepest ascent with line search | score |
| `newton_raphson()` | Second-order Newton | score, fisher |
| `bfgs()` | Quasi-Newton BFGS | score |
| `lbfgsb()` | L-BFGS-B with box constraints | score |
| `nelder_mead()` | Simplex (derivative-free) | - |
| `grid_search()` | Exhaustive grid | - |
| `random_search()` | Random sampling | - |

## Function Transformers

```{r transformers, eval=FALSE}
# Stochastic gradient (mini-batching)
loglike_sgd <- with_subsampling(loglike, data = x, subsample_size = 32)

# Regularization
loglike_l2 <- with_penalty(loglike, penalty_l2(), lambda = 0.1)
loglike_l1 <- with_penalty(loglike, penalty_l1(), lambda = 0.1)
```

## Documentation

Full documentation: <https://queelius.github.io/compositional.mle/>

## License

MIT
